---
title: "FinalProj"
author: "Adam Walters, Jijun Niu"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
nba_data <- read.csv("25-25PlayerStats.csv")

# Cleaning the data to remove outliers
# 1. Convert character columns to numeric

cols_to_fix <- c("USG.", "TS.", "TO.", "ORtg", "eFG.")
nba_data[cols_to_fix] <- lapply(nba_data[cols_to_fix], as.numeric)

# 2. Filter for play-time requirements and outliers 
cleaned_data <- subset(nba_data, 
                       MpG >= 4 & 
                       GP >= 3 & 
                       ORtg <= 160 & 
                       ORtg >= 50 & 
                       TO. >= 2 & 
                       TO. <= 50)

# 3. Remove rows with missing values in key analytical variables
critical_cols <- c("USG.", "TS.", "TO.", "ORtg")
cleaned_data <- cleaned_data[complete.cases(cleaned_data[, critical_cols]), ]

# 4. View results
summary(cleaned_data)
```


# Part 1: Multiple Linear Regression
### Implements linear regression model 
$ORtg = \beta_0 + \beta_1(USG\%) + \beta_2(TS\%) + \beta_3(TO\%) + \epsilon$

```{r}
# Multiple linear regression based on cleaned data
regression_model <- lm(ORtg ~ USG. + TS. + TO., data = cleaned_data)

# Storing summary of outputs determined via regression model
model_summary <- summary(regression_model)

# Storing the coefficient table
coef_table <- coef(model_summary)

# Storing confidence intervals
coef_intervals <- confint(regression_model)

#MULTIPLE LINEAR REGRESSION RESULTS
#Regression equation coefficients:
print(round(coef(regression_model), 4))

#Regression equation:
cat("Predicted ORtg = ",
    round(coef(regression_model)[1], 4), " + ",
    round(coef(regression_model)["USG."], 4), "(USG.) + ",
    round(coef(regression_model)["TS."], 4), "(TS.) + ",
    round(coef(regression_model)["TO."], 4), "(TO.)\n",
    sep = "")


#Coefficient table:
print(round(coef_table, 4))


#Adjusted R-squared:
cat(round(model_summary$adj.r.squared, 4), "\n")


#P-values:
print(coef_table[, "Pr(>|t|)"])


#95% confidence intervals for coefficients:
print(round(coef_intervals, 4))

# Plot model
#plot(regression_model)

par(mfrow = c(2, 2))

# Generate the 4 residual plots
plot(regression_model)


par(mfrow = c(1, 1))
```


# Part 2:Hypothesis Testing: Permutation test

```{r}
# Define groups 
high_usage <- cleaned_data$TS.[cleaned_data$USG. > 25]
low_usage <- cleaned_data$TS.[cleaned_data$USG. <= 25]

# Observed difference
obs_diff <- mean(high_usage) - mean(low_usage)

# Permutation Test 
set.seed(4714)

n_permutations <- 10000
perm_diffs <- numeric(n_permutations)
all_ts <- c(high_usage, low_usage)
n_high <- length(high_usage)

for (i in 1:n_permutations) {
  shuffled <- sample(all_ts)
  perm_high <- shuffled[1:n_high]
  perm_low <- shuffled[(n_high + 1):length(all_ts)]
  perm_diffs[i] <- mean(perm_high) - mean(perm_low)
}

# Calculate P-value (two-tailed)
p_value <- mean(abs(perm_diffs) >= abs(obs_diff))
print(paste("Permutation P-value:", p_value))

# Visualize the distribution
hist(perm_diffs, main="Distribution of Permuted Differences", 
     xlab="Difference in Mean TS%", col="skyblue")
abline(v = obs_diff, col="red", lwd=2)
```

# Part 3: Discrete Distribution (geometric)
We will model the number of shot attempts required for a player to achieve their first successful score using a geometric distribution. The success probability p will be approximated by TS%. We will
compute expected values $E[X]=1/p$ and probabilities such as $P(X>k)$to compare
consistency between players with different usage levels. Modeling number of attempts to first score using TS% as the probability of success $p$


step 1: use $E[x] = 1/p = 1/ ts%$ to find the mean value of the number of shot attempt a player need to achieve their first successful score

step 2: use geometirc distribution to find probability of the 2nd attempt be the first successful score for high usage and low usage NBS player


ctrl+ shift+ c --> comment multiple lines 
```{r}
#high usage rate
expected_value_h <- mean (1/high_usage)
expected_value_h
k  <- 2
success_less_than_2_h <- pgeom(k-1,high_usage)
hist(success_less_than_2_h, main="probability of getting the first score within 2 attempts for high usage player", xlab="Probability", col="skyblue")

# low usage rate
expected_value_l <- mean (1/low_usage)
expected_value_l

success_less_than_2_l <- pgeom(k-1,low_usage)
hist(success_less_than_2_l, main="probability of getting the first score within 2 attempts for low usage player", xlab="Probability", col="skyblue")

```

Discussion: 
High-usage players and low-usage players have a similar mean value of around 1.75 attempts to achieve their first successful score. However, the probability of scoring within the first two attempts differs between the two groups. Low-usage players show a much wider spread, with probabilities ranging from 0.30 to 1.00, while high-usage players are more concentrated between 0.70 and 0.95. This suggests that high-usage players are generally more consistent scorers



