---
title: "FinalProj"
author: "Adam Walters"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
nba_data <- read.csv("25-25PlayerStats.csv")

# Cleaning the data to remove outliers
# 1. Convert character columns to numeric

cols_to_fix <- c("USG.", "TS.", "TO.", "ORtg", "eFG.")
nba_data[cols_to_fix] <- lapply(nba_data[cols_to_fix], as.numeric)

# 2. Filter for play-time requirements and outliers 
cleaned_data <- subset(nba_data, 
                       MpG >= 4 & 
                       GP >= 3 & 
                       ORtg <= 160 & 
                       ORtg >= 50 & 
                       TO. >= 2 & 
                       TO. <= 50)

# 3. Remove rows with missing values in key analytical variables
critical_cols <- c("USG.", "TS.", "TO.", "ORtg")
cleaned_data <- cleaned_data[complete.cases(cleaned_data[, critical_cols]), ]

# 4. View results
summary(cleaned_data)
```


## Multiple Linear Regression
Implements linear regression model $ORtg = \beta_0 + \beta_1(USG\%) + \beta_2(TS\%) + \beta_3(TO\%) + \epsilon$
(revisit if we are instructed to use different method for multiple linear regression)


Hypothesis Testing: Permutation test
```{r}
# Define groups 
high_usage <- cleaned_data$TS.[cleaned_data$USG. > 25]
low_usage <- cleaned_data$TS.[cleaned_data$USG. <= 25]

# Observed difference
obs_diff <- mean(high_usage) - mean(low_usage)

# Permutation Test 
set.seed(4714)

n_permutations <- 10000
perm_diffs <- numeric(n_permutations)
all_ts <- c(high_usage, low_usage)
n_high <- length(high_usage)

for (i in 1:n_permutations) {
  shuffled <- sample(all_ts)
  perm_high <- shuffled[1:n_high]
  perm_low <- shuffled[(n_high + 1):length(all_ts)]
  perm_diffs[i] <- mean(perm_high) - mean(perm_low)
}

# Calculate P-value (two-tailed)
p_value <- mean(abs(perm_diffs) >= abs(obs_diff))
print(paste("Permutation P-value:", p_value))

# Visualize the distribution
hist(perm_diffs, main="Distribution of Permuted Differences", 
     xlab="Difference in Mean TS%", col="skyblue")
abline(v = obs_diff, col="red", lwd=2)
```

Discrete Distribution (geometric)
We will model the number of shot attempts
required for a player to achieve their first successful score using a geometric
distribution. The success probability p will be approximated by TS%. We will
compute expected values E[X]=1/p and probabilities such as P(X>k) to compare
consistency between players with different usage levels.

step 1: use E[x] = 1/p = 1/ ts%to find the mean value of the number of shot attempt a player need to achieve their first successful score
step 2: use geometirc distribution to find probability of the kth attempt be the first successful score 
step 3: using permuatation testing to c if the probability varies with differenct usage levels. 

ctrl+ shift+ c --> comment multiple lines 
```{r}
#true_shooting <- cleaned_data$TS.

#expected_value <- mean (1/true_shooting)


# k  <- 2
# success_less_than_2 <- pgeom(k-1,true_shooting)
# hist(success_less_than_2, main="probability of getting the first score within 2 attempts", 
#      xlab="miaomiao", col="skyblue")

## print expected value
## change graph label 
#high usage rate
expected_value_h <- mean (1/high_usage)
hist(expected_value_h, main="probability of getting the first score within 2 attempts high",
     xlab="miaomiao", col="skyblue")
k  <- 2
success_less_than_2_h <- pgeom(k-1,high_usage)
hist(success_less_than_2_h, main="probability of getting the first score within 2 attempts high",
     xlab="miaomiao", col="skyblue")

# low usage rate
expected_value_l <- mean (1/low_usage)
hist(expected_value_l, main="probability of getting the first score within 2 attempts high",
     xlab="miaomiao", col="skyblue")
success_less_than_2_l <- pgeom(k-1,low_usage)
hist(success_less_than_2_l, main="probability of getting the first score within 2 attempts low",
     xlab="miaomiao", col="skyblue")

```


modeling number of attempts to first score using TS% as the probability of success $p$


